Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-06-23 paper.qmd
keywords fixed and fig 5 tidied

keywords: [high-dimensional data vizualization, non-linear dimension reduction, tour]
keywords-formatted: [high-dimensional data vizualization, non-linear dimension reduction, tour]

@@

error_plot_two_curvy <- error_df_two_curvy_abs |>
  ggplot(aes(x = emb1,
             y = emb2,
             colour = error_cat)) +
  geom_point(alpha=0.5) +
  interior_annotation(label = "b", cex = 1.2) +
  scale_colour_manual(values = c("#FFFFC8", "#FFFAC0", "#FEF2B3", "#FCE9A3",
                                 "#FADE8F", "#F9D378", "#F8C65D", "#F6B938",
                                 "#F5AA00", "#F49B00", "#F28A00", "#F07800",
                                 "#ED6400", "#EA4D00", "#DE3900", "#CD2A00",
                                 "#BB1B00", "#A70C00", "#93001C", "#7D0025")) +
  xlab("tSNE1") +
  ylab("tSNE2") +
  theme_bw() +
  theme(
    aspect.ratio = 1,
    legend.position = "none"
  )

error_plot_two_curvy_quasi <- ggplot(
  error_df_two_curvy_abs, 
  aes(x = sqrt_row_wise_total_error, 
      y = 0, 
      colour = error_cat)) +
  geom_quasirandom(alpha = 0.5) +
  interior_annotation(label = "a", cex = 1.2) +
  scale_colour_manual(values = c("#FFFFC8", "#FFFAC0", "#FEF2B3", "#FCE9A3",
                                 "#FADE8F", "#F9D378", "#F8C65D", "#F6B938",
                                 "#F5AA00", "#F49B00", "#F28A00", "#F07800",
                                 "#ED6400", "#EA4D00", "#DE3900", "#CD2A00",
                                 "#BB1B00", "#A70C00", "#93001C", "#7D0025")) +
  xlab(expression(e[hj])) +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position = "none",
        plot.margin = margin(0, 0, 0, 0))
```

@@

two_curvy_proj_tsne_first_model1_coloured <- projected_df |>
  ggplot(
    aes(
      x = proj1,
      y = proj2,
      colour = cluster)) +
  geom_point(
    size = 0.5,
    alpha = 0.2) +
  geom_segment(
    data=axes,
    aes(x=x1, y=y1, xend=x2, yend=y2),
    colour="grey70") +
  geom_text(
    data=axes,
    aes(x=x2, y=y2),
    label=rownames(axes),
    colour="grey50",
    size = 2) +
  geom_path(
    data=circle,
    aes(x=c1, y=c2), colour="grey70") +
  scale_colour_manual(values = c("#FFFFC8", "#FFFAC0", "#FEF2B3", "#FCE9A3",
                                 "#FADE8F", "#F9D378", "#F8C65D", "#F6B938",
                                 "#F5AA00", "#F49B00", "#F28A00", "#F07800",
                                 "#ED6400", "#EA4D00", "#DE3900", "#CD2A00",
                                 "#BB1B00", "#A70C00", "#93001C", "#7D0025")) +
  xlim(c(-0.6, 0.6)) +
  ylim(c(-0.6, 0.6)) +
  interior_annotation(label = "c", cex = 1.2) +
  theme(aspect.ratio = 1,
        legend.position = "none")

@@

```{r}
#| label: fig-p-d-error-in-2d-two-curvy
#| fig-cap: "Explore the distribution of residuals in \\gD{} and \\fD{}. Color indicates error ($e_{hj}$), dark color indicating high error and light indicates low error. Most large errors are distributed in one cluster (bottom one in c) and most small errors are distributed in the other cluster."
#| out-width: 80%
#| fig-pos: H

>>>>

keywords: [high-dimensional data, data visualization, tour, statistical graphics, exploratory data analysis, unsupervised learning]
keywords-formatted: [high-dimensional data, data visualization, tour, statistical graphics, exploratory data analysis, unsupervised learning]

@@

error_plot_two_curvy <- error_df_two_curvy_abs |>
  ggplot(aes(x = emb1,
             y = emb2,
             colour = error_cat)) +
  geom_point(alpha=0.5) +
  interior_annotation(label = "b", cex = 1.5) +
  scale_colour_manual(values = c("#FFFFC8", "#FFFAC0", "#FEF2B3", "#FCE9A3",
                                 "#FADE8F", "#F9D378", "#F8C65D", "#F6B938",
                                 "#F5AA00", "#F49B00", "#F28A00", "#F07800",
                                 "#ED6400", "#EA4D00", "#DE3900", "#CD2A00",
                                 "#BB1B00", "#A70C00", "#93001C", "#7D0025")) +
  xlab("tSNE1") +
  ylab("tSNE2") +
  theme_bw() +
  theme(
    aspect.ratio = 1,
    legend.position = "none",
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank()
  )

error_plot_two_curvy_quasi <- ggplot(
  error_df_two_curvy_abs, 
  aes(x = sqrt_row_wise_total_error, 
      y = 0, 
      colour = error_cat)) +
  geom_quasirandom(alpha = 0.5) +
  interior_annotation(label = "a", cex = 1.5) +
  scale_colour_manual(values = c("#FFFFC8", "#FFFAC0", "#FEF2B3", "#FCE9A3",
                                 "#FADE8F", "#F9D378", "#F8C65D", "#F6B938",
                                 "#F5AA00", "#F49B00", "#F28A00", "#F07800",
                                 "#ED6400", "#EA4D00", "#DE3900", "#CD2A00",
                                 "#BB1B00", "#A70C00", "#93001C", "#7D0025")) +
  xlab(expression(e[hj])) +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position = "none",
        plot.margin = margin(0, 0, 0, 0))
```

@@

two_curvy_proj_tsne_first_model1_coloured <- projected_df |>
  ggplot(
    aes(
      x = proj1,
      y = proj2,
      colour = cluster)) +
  geom_point(
    size = 0.5,
    alpha = 0.2) +
  geom_segment(
    data=axes,
    aes(x=x1, y=y1, xend=x2, yend=y2),
    colour="grey70") +
  geom_text(
    data=axes,
    aes(x=x2, y=y2),
    label=rownames(axes),
    colour="grey50",
    size = 2) +
  geom_path(
    data=circle,
    aes(x=c1, y=c2), colour="grey70") +
  scale_colour_manual(values = c("#FFFFC8", "#FFFAC0", "#FEF2B3", "#FCE9A3",
                                 "#FADE8F", "#F9D378", "#F8C65D", "#F6B938",
                                 "#F5AA00", "#F49B00", "#F28A00", "#F07800",
                                 "#ED6400", "#EA4D00", "#DE3900", "#CD2A00",
                                 "#BB1B00", "#A70C00", "#93001C", "#7D0025")) +
  xlim(c(-0.6, 0.6)) +
  ylim(c(-0.6, 0.6)) +
  interior_annotation(label = "c", cex = 1.5) +
  theme(aspect.ratio = 1,
        legend.position = "none")

@@

```{r}
#| label: fig-p-d-error-in-2d-two-curvy
#| fig-cap: "Examining the distribution of residuals in a jittered dotplot (a), \\gD{} NLDR layout (b) and a tour of \\fD{} data space (c). Color indicates error ($e_{hj}$), dark color indicating high error and light indicates low error. Most large errors are distributed in one cluster (bottom one in c) and most small errors are distributed in the other cluster."
#| out-width: 100%
#| fig-pos: H
#| fig-width: 8
#| fig-height: 3
